---
title: 'smoothSDE Analysis of Atlantic BRS Ziphius DTAG Data: smoothSDEs'
author: "Stacy DeRuiter"
date: "2023-2024"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# note: must install Theo Michelot's smooth SDE package first
# remotes::install_github('TheoMichelot/smoothSDE')
library(tidyverse)
library(ggformula)
library(smoothSDE)
library(cowplot)
library(R.matlab)
library(extrafont)

theme_set(theme_bw(base_size = 16) + theme(text = element_text(family="Times New Roman")))
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = 'figures/zc-sde-',
                      dev = c('jpeg', 'png'),
                      dpi = 600,
                      fig.width = 5, fig.height = 3.5)
```

# Background

## Methods Paragraph

Whale depth and heading data from the DTAGs were decimated to 1/3 Hz (after first applying a no-delay low-pass anti-alias filter). 

- *Option: For depth, we computed the first difference of the depth measurements to reduce temporal autocorrelation in the time-series before modeling.*

Dives to greater than 50 meters depth were detected, the classified as deep, presumed foraging dives (dives to over 800m depth) or shallower dives (less than 800m) (Shearer et. al. ref). Each dive was also classified as exposed if there were any controlled MFAS exposure sounds during that dive, and baseline if it occurred before any controlled exposure. We fitted four varying-coefficient stochastic differential equations (SDEs) (Michelot et al. 2022, 2023): for heading and depth-difference data, in shallow and deep dives respectively. Each model describes the heading or depth as a function of proportion of dive-time elapsed, and employs a difference smooth to quantify differences between baseline and exposed dives. The model also includes a random effect to account for whale-to-whale differences, which also allows for visualization of estimated individual responses.

Data processing used the package tagtools (DeRuiter et al. 2023), and models were fitted using the package smoothSDE (Michelot 2021, 2022, 2023).

DeRuiter S, Johnson M, Sweeney D, McNamara-Oh Y, Fynewever S, Tejevbo O,
  Marques T, Wang Y, Ogedegbe O. (2023). tagtools: Work with Data from
  High-Resolution Biologging Tags. R package version 0.1.0,
  https://CRAN.R-project.org/package=tagtools.
  
Michelot, T., Glennie, R., Harris, C., Thomas, L. (2021)
"Varying-coefficient stochastic differential equations with applications in ecology"
Journal of Agricultural, Biological, and Environmental Statistics. DOI: 10.1007/s13253-021-00450-6.

Michelot, Théo, et al. "Continuous-time modelling of behavioural responses in animal movement." arXiv preprint arXiv:2212.09574 (2022).

Michelot T (2023). smoothSDE: Varying-coefficient stochastic differential
  equations. R package version 0.1.

# Preparation

## Read in and prepare data

The data has undergone additional pre-processing in Matlab. See files "prep-SDEs.Rmd" and "smoothSDE_prep.m" for details.

Data from a given dive will be considered exposed to the CEE if any part of that dive overlapped with the CEE.

The real MFAS dataset is included in  the data prep steps and initial graphs, but is not analyzed with the smoothSDE model for now.

```{r}
abrs_zc <- read_csv('data/AtlBRS_Zc_smoothSDE_data.csv',
                    show_col_types = FALSE) |>
  # filter out times before/after first/last dive
  drop_na(diveprop, max_depth) |>
  # filter out post-dive surface intervals
  dplyr::filter(diveprop <= 1) |>
  # append "dive-" to dive ID numbers
  mutate(time = time_hr # we want to keep hr for clarity of units but smoothSDE requires time var
         ) |> 
  rename(depth = p_lo,
         pitch = pitch_lo,
         roll = roll_lo,
         heading = head_lo_deg
         ) |>
  # re-create new ID variable so each whale-dive has a unique identifier
  mutate(ID = paste0('dive-', as.numeric(interaction(whaleID, ID, drop = TRUE))))
```

## CEEs

Add in information on CEEs. 

Since audits are not available for all tag deployments, this info is not all easily accessible directly from the data files. Note also some of the tag deployment dates are wrong on quicklook doc files, so the dates used here have been validated by checking the Julian-day numbers in the tag ID string and, where needed, the .xml cal files.

Info used is:

```{r}
# cee_meta_from_will <- read_csv('https://raw.githubusercontent.com/atlanticbrs/zcss_data_dev/main/00_data_input/cee/cee_metadata_flat.csv?token=GHSAT0AAAAAACOPSCEQWIEZL3QN6XLNQQXUZO5FVYQ')
CEE_meta <- tibble(whaleID = c("zc17_234a", 
                               "zc19_218a",
                               "zc20_232a", 
                               "zc22_219a"),
                   tagon_time = lubridate::ymd_hms('2017 aug 22 15:10:46',
                                               '2019 aug 6 15:38:09',
                                               '2020 aug 19 15:28:50',
                                               '2022 aug 7 13:56:55'),
                   # based on cee_metadata_flat
                   cee_start_time = lubridate::ymd_hms('2017 aug 22 18:41:00',
                                               '2019 aug 6 17:37:43',
                                               '2020 aug 19 17:43:23',
                                               '2022 aug 7 17:09:00'),
                   cee_end_time = lubridate::ymd_hms('2017 aug 22 18:54:00',
                                               '2019 aug 6 18:07:43',
                                               '2020 aug 19 18:13:22',
                                               '2022 aug 7 18:10:00'),
                   cee_type = c('scaled MFAS', 
                                'scaled MFAS', 
                                'scaled MFAS', 
                                'real MFAS')
)
```

Note that the 2017 whale only did one long and deep dive, which ended before the start of the CEE. So if this analysis is looking at deep dives only, then this whale has "no exposed" deep dives.

```{r}
CEE_meta <- CEE_meta |>
  mutate(
    cee_start_hr = round(as.numeric(
      difftime(cee_start_time, tagon_time,
               units = 'hours')
    ),
    digits = 3),
    cee_end_hr = round(as.numeric(
      difftime(cee_end_time, tagon_time,
               units = 'hours')
    ),
    digits = 3)
  )

knitr::kable(CEE_meta)
```

```{r, message = FALSE}
abrs_zc <- left_join(abrs_zc, CEE_meta, by = 'whaleID') |>
  mutate(cee_status = case_when( time_hr < cee_start_hr ~ 'pre',
                                 time_hr >= cee_start_hr & time_hr <= cee_end_hr ~ 'during',
                                 time_hr > cee_end_hr ~ 'post'),
         cee_status = fct_relevel(cee_status, 'pre', 'during', 'post')) |>
  group_by(whaleID, ID) |>
  # this one classes each entire DIVE as during if any part of it was during
  mutate(cee_dive_status_pooled = ifelse(sum(time_hr >= cee_start_hr & time_hr <= cee_end_hr) > 1,
                                  'CEE',
                                  'pre or post'),
         cee_dive_status = ifelse(cee_dive_status_pooled == "CEE", 
                                  paste0(whaleID, cee_dive_status_pooled),
                                  cee_dive_status_pooled)) |>
  ungroup()
```

## Echolocation

Add in information on the times of echolocation clicks.

```{r}
whales <- unique(pull(abrs_zc, whaleID))
echolocation <- list()

for (w in c(1:length(whales))){
  click_files <- dir(file.path('data/echolocation', whales[w]))
  click_files <- file.path('data/echolocation', whales[w], click_files)
  for (f in c(1:length(click_files))){
    these_clicks <- R.matlab::readMat(click_files[f])
    # note: 1st list-item from readMat() is the data vector/matrix, later items are attributes. 
    # the NAME of the data is different from file to file hence use of numeric index
    if (f == 1){
      echolocation[[w]] <- tibble(whaleID = whales[w],
                                  soc = min(these_clicks[[1]], na.rm = TRUE),
                                  eoc = max(these_clicks[[1]], na.rm = TRUE))
    }else{
    echolocation[[w]] <- bind_rows(echolocation[[w]], 
                                  tibble(whaleID = whales[w],
                                         soc = min(these_clicks[[1]], na.rm = TRUE),
                                         eoc = max(these_clicks[[1]], na.rm = TRUE)))
    }
  }
}

echolocation <- bind_rows(echolocation) |>
  # get soc and eoc in hours
  mutate(soc_hr = soc / 3600,
         eoc_hr = eoc / 3600)

# next need to: put a binary variable clicking/not into the time series data for plotting
abrs_zc <- mutate(abrs_zc, clicking = 0, foraging_period = 0)
for (d in c(1:nrow(echolocation))){
  abrs_zc <- abrs_zc |>
    mutate(clicking = if_else(whaleID == pull(echolocation, whaleID)[d] & 
                                  time_hr >= pull(echolocation, soc_hr)[d] & 
                                  time_hr <= pull(echolocation, eoc_hr)[d],
                              1,
                              clicking),
           foraging_period = if_else(whaleID == pull(echolocation, whaleID)[d] & 
                                  time_hr >= pull(echolocation, soc_hr)[d] & 
                                  time_hr <= pull(echolocation, eoc_hr)[d],
                              d,
                              foraging_period)) 
}

abrs_zc <- mutate(abrs_zc,
                  soc = if_else(clicking - lag(clicking) == 1, 1, 0),
                  eoc = if_else(clicking - lag(clicking) == -1, 1, 0))
```


## Data Exploration/Verification

There was some question at some point if the CEE start/end times were correct, so these plots are shown so that the researchers who know the datasets well can verify that the assignment of pre/during/post exposure status is correct.

```{r, cee-status-exact, fig.height=7.5, fig.width = 6.5}
gf_line(depth ~ time_hr, 
        color = ~cee_status,
        data = abrs_zc) |>
  gf_path(depth ~ time_hr, 
        color = ~cee_status,
        group = ~foraging_period,
        data = abrs_zc |> filter(clicking == 1),
        linewidth = 1.25) |>
  gf_facet_wrap(~whaleID, scales = 'free_x', nrow = 4) |>
  gf_theme(scale_color_brewer('CEE', palette = 'Dark2')) |>
  gf_lims(y = c(1800, 0)) |>
  gf_labs(x = 'Time (h since tag start)', y = 'Depth (m)')
```

At this point we can see that the last shallow dives by zc20_232a and zc19_218a both have the tag fall off before the end of the dive. So these two shallow dives should be removed from analysis.

```{r}
abrs_zc <- abrs_zc |>
  group_by(whaleID) |>
  filter(whaleID == "zc17_234a" |
         (whaleID == "zc19_218a" & ID != last(ID)) |
         (whaleID == "zc20_232a" & ID != last(ID)) |
         (whaleID == "zc22_219a")
         ) |>
  ungroup()
```

```{r, cee-status-exact-complete, fig.height=7.5, fig.width = 6.5}
gf_line(depth ~ time_hr, 
        color = ~cee_status,
        data = abrs_zc) |>
  gf_path(depth ~ time_hr, 
        color = ~cee_status,
        group = ~foraging_period,
        data = abrs_zc |> filter(clicking == 1),
        linewidth = 1.25) |>
  gf_facet_wrap(~whaleID, scales = 'free_x', nrow = 4) |>
  gf_theme(scale_color_brewer('CEE', palette = 'Dark2')) |>
  gf_lims(y = c(1800, 0)) |>
  gf_labs(x = 'Time (h since tag start)', y = 'Depth (m)')
```

```{r, cee-status-by-dive, fig.height=7.5, fig.width = 6.5}
gf_line(depth ~ time_hr, 
        color = ~cee_dive_status,
        group = ~1,
        data = abrs_zc) |>
  gf_path(depth ~ time_hr, 
        color = ~cee_dive_status,
        group = ~foraging_period,
        data = abrs_zc |> filter(clicking == 1),
        linewidth = 1.25) |>
  gf_facet_wrap(~whaleID, scales = 'free_x', nrow = 4) |>
  gf_theme(scale_color_brewer('CEE', palette = 'Set2')) |>
  gf_lims(y = c(1800, 0)) |>
  gf_labs(x = 'Time (h since tag start)', y = 'Depth (m)') 
```

## Split Deep and Shallow

We divide the data into two separate datasets, with deep and shallow dives. For shallow dives, discard all post-exposure dives (they are not to be considered baseline). For deep dives, post-exposure dives are also effectively discarded because there were no post-exposure deep dives in the three whales under study here.

```{r}
abrs_zc_deep <- abrs_zc |>
  filter(max_depth >= 800)

abrs_zc_shallow <- abrs_zc |>
  filter(max_depth < 800 & cee_status != "post")
```

```{r, cee-status-by-dive-deep, fig.height=7.5, fig.width = 6.5, warning = FALSE}
gf_line(depth ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = abrs_zc_deep) |>
  gf_point(depth ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = abrs_zc_deep |> filter(soc == 1 | eoc == 1),
        size = 2) |>
  gf_facet_wrap(~whaleID, scales = 'free_x', nrow = 4) |>
  gf_theme(scale_color_brewer('CEE', palette = 'Set2')) |>
  gf_lims(y = c(1800, 0)) |>
  gf_labs(x = 'Proportion total dive time', y = 'Depth (m)',
          title = 'Deep Dives')
```


Note that almost all of the non-exposed deep dives are post-exposure dives in the zc_22 tagout with the real MFAS exposure. The other three tagouts each have one pre-exposure dive.

```{r, cee-status-by-dive-shallow, fig.height=7.5, fig.width = 6.5, warning = FALSE}
gf_line(depth ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = abrs_zc_shallow) |>
  gf_facet_wrap(~whaleID, scales = 'free_x', nrow = 4) |>
  gf_theme(scale_color_brewer('CEE', palette = 'Set2')) |>
  gf_lims(y = c(800, 0)) |>
  gf_labs(x = 'Proportion total dive time', y = 'Depth (m)',
          title = 'Shallow Dives') 
```
## Final data prep

The real MFAS deployment, zc22_219a, is removed from analysis for now.

Before fitting a model with the `smoothSDE` package we must also make sure that the data is a `data.frame()` (not a `tibble()`) and that all categorical variables are factors, not character.

```{r}
abrs_zc_deep <- abrs_zc_deep |>
  dplyr::filter(whaleID != 'zc22_219a') |>
  group_by(whaleID, ID) |>
  mutate(diff_depth = c(0, diff(depth)),
         head_rad = heading / 180 * pi,
         diff_head = atan2(sin(head_rad - lag(head_rad, n = 1)),
                           cos(head_rad - lag(head_rad, n = 1)))) |>
  ungroup() |>
  dplyr::select(whaleID, ID, time, diveprop, 
                cee_dive_status, depth, diff_depth, heading, diff_head,
                soc, eoc, clicking) |>
  as.data.frame() |>
  mutate(across(where(is_character), as_factor)) |>
  mutate(cee_dive_status = factor(cee_dive_status, ordered = TRUE))

abrs_zc_shallow <- abrs_zc_shallow |>
  dplyr::filter(whaleID != 'zc22_219a') |>
  group_by(whaleID, ID) |>
  mutate(diff_depth = c(0, diff(depth)),
         head_rad = heading / 180 * pi,
         diff_head = atan2(sin(head_rad - lag(head_rad, n = 1)),
                           cos(head_rad - lag(head_rad, n = 1)))) |>
  ungroup() |>
  dplyr::select(whaleID, ID, time, diveprop, cee_dive_status, depth, diff_depth, heading, diff_head) |>
  as.data.frame() |>
  mutate(across(where(is_character), as_factor)) |>
  mutate(cee_dive_status = factor(cee_dive_status, ordered = TRUE))
```

```{r, head-by-dive-deep, fig.height=7.5, fig.width = 6.5, warning = FALSE}
gf_line(heading ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = abrs_zc_deep) |>
  gf_facet_wrap(~whaleID, scales = 'free_x', nrow = 4) |>
  gf_theme(scale_color_brewer('CEE', palette = 'Set2')) |>
  gf_labs(x = 'Proportion total dive time', y = 'Heading (degrees)',
          title = 'Deep Dives')

gf_line(diff_head / pi * 180 ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = abrs_zc_deep) |>
  gf_facet_wrap(~whaleID, scales = 'free_x', nrow = 4) |>
  gf_theme(scale_color_brewer('CEE', palette = 'Set2')) |>
  gf_labs(x = 'Proportion total dive time', y = 'Heading Difference (degrees)',
          title = 'Deep Dives')
```

```{r, head-by-dive-shal, fig.height=7.5, fig.width = 6.5, warning = FALSE}
gf_line(heading ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = abrs_zc_shallow) |>
  gf_facet_wrap(~whaleID, scales = 'free_x', nrow = 4) |>
  gf_theme(scale_color_brewer('CEE', palette = 'Set2')) |>
  gf_labs(x = 'Proportion total dive time', y = 'Heading (degrees)',
          title = 'Shallow Dives')

gf_line(diff_head / pi * 180 ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = abrs_zc_shallow) |>
  gf_facet_wrap(~whaleID, scales = 'free_x', nrow = 4) |>
  gf_theme(scale_color_brewer('CEE', palette = 'Set2')) |>
  gf_labs(x = 'Proportion total dive time', y = 'Heading Difference(degrees)',
          title = 'Shallow Dives')
```

# Models: Deep Dives
## Depth data

To reduce the amount of temporal autocorrelation in the time series data, we model the first difference of  the depths (`diff_depth`) instead of the original depth data.

We could include a difference smooth either on the drift (directionality) or on the diffusion (variability) parameter for the depth data. 

Here, we include the response on the diffusion parameter.

```{r, deep-depth-model}
r1file <- 'models/deep_ddepth.RDS'
if (file.exists(r1file)){
  deep_ddepth_sde <- readRDS(r1file)
}else{
  f_deep_ddepth <- list(mu = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts"),
           sigma = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts"))
  # Create and fit SDE
  deep_ddepth_sde <- SDE$new(formulas = f_deep_ddepth,
                             data = abrs_zc_deep, 
                             type = "BM", 
                             response = "diff_depth")
  deep_ddepth_sde$fit(silent = TRUE)
  saveRDS(deep_ddepth_sde, r1file)
}
```

```{r, deep-depth-t-model}
r1file <- 'models/deep_ddeptht.RDS'
if (file.exists(r1file)){
  deep_ddeptht_sde <- readRDS(r1file)
}else{
  f_deep_ddeptht <- list(mu = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts"),
           sigma = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts"))
  # Create and fit SDE
  deep_ddeptht_sde <- SDE$new(formulas = f_deep_ddeptht,
                             data = abrs_zc_deep, 
                             type = "BM_t", 
                             other_data = list(df = 3),
                             response = "diff_depth")
  deep_ddeptht_sde$fit(silent = TRUE)
  saveRDS(deep_ddeptht_sde, r1file)
}
```

```{r, deep-depth-plot}
deep_ddepth_sde$plot_par("diveprop")
deep_ddepth_sde$plot_par("ID")
```

We can plot the difference smooth for $\sigma$ on the linear predictor scale,
<!-- what if we want natural scale?? -->
together with 95% confidence intervals, as follows.

These plots show the change in the diffusion parameter sigma during CEE compared to baseline. Sigma measures the diffusion or variability of the response variable (either depth difference or heading). So a positive change means additional variation and negative mean less variability. 

The numeric values for change in sigma are on the scale of the linear predictor, so can be thought of as being in some kind of relative units (note I don't have time/tools to present these on the response scale of meters and degrees instead - I'm not sure enough of the model equation and CI determination code being used to be sure I can just apply an inverse link function to the displayed values).

```{r, deep-depth-diffsmooth-zc19_218a}
# Plot difference smooth on sigma for exposed dives

deep_ddeptht_sde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = "zc19_218aCEE"), term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = 'zc19_218', y = 'Depth Rate\nDifference Smooth', x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)
```

```{r, deep-depth-diffsmooth-zc20_232a}
deep_ddeptht_sde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = "zc20_232aCEE"), term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = 'zc20_232', y = 'Depth Rate\nDifference Smooth', x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)
```


## Heading data

Should we be modeling heading? Or difference in heading? Michelot et al just used angles with no worries...but perhaps they got lucky. In our case, either one (heading or diff(heading)) includes many "circular moves" (large changes and/or heading going from for example 350 to 10 degrees).

```{r, deep-head-model}
deep_head_sde_file <- 'models/deep_head.RDS'
if (file.exists(deep_head_sde_file)){
  deep_head_sde <- readRDS(deep_head_sde_file)
}else{
  deep_head_f <- list(mu = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts"),
             sigma = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts"))
  # Create and fit SDE
  deep_head_sde <- SDE$new(formulas = deep_head_f, 
                  data = abrs_zc_deep, 
                  type = "BM", 
                  response = "heading")
  deep_head_sde$fit(silent = TRUE)
  saveRDS(deep_head_sde, deep_head_sde_file)
}
```

```{r, deep-head-t-model}
deep_head_tsde_file <- 'models/deep_head_t.RDS'
if (file.exists(deep_head_tsde_file)){
  deep_head_tsde <- readRDS(deep_head_tsde_file)
}else{
  deep_head_f <- list(mu = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts")
             ,
             sigma = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts")
             )
  # Create and fit SDE
  deep_head_tsde <- SDE$new(formulas = deep_head_f, 
                  data = abrs_zc_deep, 
                  type = "BM_t", 
                  response = "heading",
                  other_data = list(df = 3)
                  )
  deep_head_tsde$fit(silent = TRUE)
  saveRDS(deep_head_tsde, deep_head_tsde_file)
}
```


```{r, deep-head-t-diffsmooth-zc19_218a}
# Plot difference smooth on sigma for exposed dives

deep_head_tsde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = "zc19_218aCEE"), term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = 'zc19_218', y = 'Change in Diffusion Parameter\n(Deep Heading)', x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)
```


```{r, deep-head-t-diffsmooth-zc20_232a}
deep_head_tsde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = "zc20_232aCEE"), term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = 'zc20_232', y = 'Change in Diffusion Parameter\n(Deep Heading)', x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)
```

### Model validation for deep models

#### Residual distribution

```{r, prep-deep-residuals}
residz <- sapply(list(deep_ddepth_sde,
                      deep_ddeptht_sde,
                      deep_head_sde,
                      deep_head_tsde
                      ), 
                 function(sde) sde$residuals()) |>
  as.data.frame()

names(residz) <- c('Depth difference',
                   'Depth difference (t)',
                   'Heading', 
                   'Heading (t)'
                   )

residz <- residz |>
  mutate(ID = pull(abrs_zc_deep, ID))

residz <- pivot_longer(residz, cols = `Depth difference`:`Heading (t)`, 
                       names_to = 'variable', 
                       values_to = 'residuals')
```

```{r, resid-hist-deep-depthdiff, fig.height=3, fig.width=5}
gf_dhistogram(~residuals, data = residz |>
                filter(variable == "Depth difference")) |>
  gf_dist(dist = 'norm',
          # df = 3, 
          color = 'grey44') |>
  # here t would be WORSE
  # gf_dist(dist = 't',
  #         df = 3, 
  #         color = 'grey77') |>
  gf_lims(x = c(-5, 5)) |>
  gf_labs(title = 'Deep Depth Difference')
```

```{r, resid-hist-deep-depthdifft, fig.height=3, fig.width=5}
gf_dhistogram(~residuals, data = residz |>
                filter(variable == "Depth difference (t)")) |>
  gf_dist(dist = 't',
          df = 3,
          color = 'grey44') |>
  gf_lims(x = c(-5, 5)) |>
  gf_labs(title = 'Deep Depth Difference')
```

```{r, resid-hist-deep-head, fig.height=3, fig.width=5}
gf_dhistogram(~residuals, data = residz |>
                filter(variable == "Heading")) |>
  gf_dist(dist = 'norm',
          color = 'grey44') |>
  gf_lims(x = c(-5, 5)) |>
  gf_labs(title = 'Deep Heading')
```

```{r, resid-hist-deep-head-t, fig.height=3, fig.width=5}
gf_dhistogram(~residuals, data = residz |>
                filter(variable == "Heading (t)")) |>
  gf_dist(dist = 't',
          df = 3,
          color = 'grey44') |>
  gf_lims(x = c(-5, 5)) |>
  gf_labs(title = 'Deep Heading (t)')
```

Here, the t model for the heading data looks better, and the normal model for the depth difference model looks ok.

#### Residual independence

```{r, resid-acfs-deep, fig.width = 5, fig.height = 3}
CalvinBayes::acf_plot(filter(residz,
                             variable == "Depth difference") |>
                        pull(residuals) |>
                        na.omit()) |>
  gf_labs(title = "Depth Difference", y = 'Residual ACF', x = 'Lag')

CalvinBayes::acf_plot(filter(residz,
                             variable == "Depth difference (t)") |>
                        pull(residuals) |>
                        na.omit()) |>
  gf_labs(title = "Depth Difference (t)", y = 'Residual ACF', x = 'Lag')

CalvinBayes::acf_plot(filter(residz,
                             variable == "Heading") |>
                        pull(residuals) |>
                        na.omit()) |>
  gf_labs(title = "Heading", y = 'Residual ACF', x = 'Lag')

CalvinBayes::acf_plot(filter(residz,
                             variable == "Heading (t)") |>
                        pull(residuals) |>
                        na.omit()) |>
  gf_labs(title = "Heading (t)", y = 'Residual ACF', x = 'Lag')
```

Residuals of *all* the models for the deep dives indicate that there are remaining, un-modeled temporal patterns in the data.

# Models: Shallow Dives
## Depth data

To reduce the amount of temporal autocorrelation in the time series data, we model the first difference of  the depths (`diff_depth`) instead of the original depth data.

We could include a difference smooth either on the drift (directionality) or on the diffusion (variability) parameter for the depth data. 

Here, we include the response on the diffusion parameter.

```{r}
r1file <- 'models/shallow_ddepth.RDS'
if (file.exists(r1file)){
  shallow_ddepth_sde <- readRDS(r1file)
}else{
  f_shallow_ddepth <- list(mu = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts"),
           sigma = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts"))
  # Create and fit SDE
  shallow_ddepth_sde <- SDE$new(formulas = f_shallow_ddepth,
                             data = na.omit(abrs_zc_shallow), 
                             type = "BM", 
                             response = "diff_depth")
  shallow_ddepth_sde$fit(silent = TRUE)
  saveRDS(shallow_ddepth_sde, r1file)
}
```

```{r}
r1file <- 'models/shallow_ddeptht.RDS'
if (file.exists(r1file)){
  shallow_ddeptht_sde <- readRDS(r1file)
}else{
  f_shallow_ddeptht <- list(mu = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts"),
           sigma = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts"))
  # Create and fit SDE
  shallow_ddeptht_sde <- SDE$new(formulas = f_shallow_ddeptht,
                             data = na.omit(abrs_zc_shallow), 
                             type = "BM_t", 
                             other_data = list(df = 3),
                             response = "diff_depth")
  shallow_ddeptht_sde$fit(silent = TRUE)
  saveRDS(shallow_ddeptht_sde, r1file)
}
```

```{r}
shallow_ddepth_sde$plot_par("diveprop")
shallow_ddepth_sde$plot_par("ID")
```

We can plot the difference smooth for $\sigma$ on the linear predictor scale,
<!-- what if we want natural scale?? -->
together with 95% confidence intervals, as follows.

```{r, shallow-depth-diffsmooth-zc17_234a}
# Plot difference smooth on sigma for exposed dives
shallow_ddepth_sde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = "zc17_234aCEE"), term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = 'zc17_234', y = 'Depth Rate\nDifference Smooth', x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)
```

```{r, shallow-depth-diffsmooth-zc20_232a}
shallow_ddepth_sde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = "zc20_232aCEE"), term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = 'zc20_232', y = 'Depth Rate\nDifference Smooth', x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)
```


## Heading data

Same caveats discussed earlier - about modeling circular data with a non-circular model - apply for shallow dive models as for the deep dive models.

```{r, shallow-head-model}
shallow_head_sde_file <- 'models/shallow_head.RDS'
if (file.exists(shallow_head_sde_file)){
  shallow_head_sde <- readRDS(shallow_head_sde_file)
}else{
  shallow_head_f <- list(mu = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts"),
             sigma = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts"))
  # Create and fit SDE
  shallow_head_sde <- SDE$new(formulas = shallow_head_f, 
                  data = abrs_zc_shallow, 
                  type = "BM", 
                  response = "heading")
  shallow_head_sde$fit(silent = TRUE)
  saveRDS(shallow_head_sde, shallow_head_sde_file)
}
```

```{r, shallow-head-t-model}
shallow_head_tsde_file <- 'models/shallow_head_t.RDS'
if (file.exists(shallow_head_tsde_file)){
  shallow_head_tsde <- readRDS(shallow_head_tsde_file)
}else{
  shallow_head_f <- list(mu = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts")
             ,
             sigma = ~ s(ID, bs = "re") +
               s(diveprop, k = 10, bs = "ts") +
               s(diveprop, by = cee_dive_status, k = 10, bs = "ts")
             )
  # Create and fit SDE
  shallow_head_tsde <- SDE$new(formulas = shallow_head_f, 
                  data = abrs_zc_shallow, 
                  type = "BM_t", 
                  response = "heading",
                  other_data = list(df = 3)
                  )
  shallow_head_tsde$fit(silent = TRUE)
  saveRDS(shallow_head_tsde, shallow_head_tsde_file)
}
```


```{r, shallow-head-t-diffsmooth-zc17_234a}
# Plot difference smooth on sigma for exposed dives

shallow_head_tsde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = "zc17_234aCEE"), term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = 'zc17_234', y = 'Change in Diffusion Parameter\n(Shallow Heading)', x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)
```

```{r, shallow-head-t-diffsmooth-zc20_232a}
shallow_head_tsde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = "zc20_232aCEE"), term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = 'zc20_232', y = 'Change in Diffusion Parameter\n(Shallow Heading)', x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)
```

### Model validation for shallow models

```{r}
resids_shal_ddepth <- shallow_ddepth_sde$residuals()
resids_shal_ddeptht <- shallow_ddeptht_sde$residuals()
resids_shal_head <- shallow_head_sde$residuals()
resids_shal_headt <- shallow_head_tsde$residuals()
```

```{r, resid-hist-shallow-depthdiff, fig.height=3, fig.width=5}
gf_dhistogram(~resids_shal_ddepth) |>
  gf_dist(dist = 'norm',
          color = 'grey44') |>
  gf_lims(x = c(-5, 5)) |>
  gf_labs(title = 'Shallow Depth Difference')
```

```{r, resid-hist-shallow-depthdifft, fig.height=3, fig.width=5}
gf_dhistogram(~resids_shal_ddeptht) |>
  gf_dist(dist = 'norm',
          color = 'grey44') |>
  gf_lims(x = c(-5, 5)) |>
  gf_labs(title = 'Shallow Depth Difference (t)')
```

```{r, resid-hist-shallow-head, fig.height=3, fig.width=5}
gf_dhistogram(~resids_shal_head) |>
  gf_dist(dist = 'norm',
          color = 'grey44') |>
  gf_lims(x = c(-5, 5)) |>
  gf_labs(title = 'Shallow Heading')
```

```{r, resid-hist-shallow-head-t, fig.height=3, fig.width=5}
gf_dhistogram(~resids_shal_headt) |>
  gf_dist(dist = 't',
          df = 3,
          color = 'grey77') |>
  gf_lims(x = c(-5, 5)) |>
  gf_labs(title = 'Shallow Heading')
```

```{r, resid-acfs-shal, fig.width = 5, fig.height = 3}
CalvinBayes::acf_plot(resids_shal_ddepth |>
                        na.omit()) |>
  gf_labs(title = "Depth Difference", y = 'Residual ACF', x = 'Lag')

CalvinBayes::acf_plot(resids_shal_ddeptht |>
                        na.omit()) |>
  gf_labs(title = "Depth Difference (t)", y = 'Residual ACF', x = 'Lag')

CalvinBayes::acf_plot(resids_shal_head |>
                        na.omit()) |>
  gf_labs(title = "Heading", y = 'Residual ACF', x = 'Lag')

CalvinBayes::acf_plot(resids_shal_headt |>
                        na.omit()) |>
  gf_labs(title = "Heading (t)", y = 'Residual ACF', x = 'Lag')
```

# Figures for paper


```{r, zc19_218a-panels, fig.width = 7.5, fig.height = 3}
colrz <- RColorBrewer::brewer.pal(3, 'Set2')
colrz[1] <- colrz[3] <- 'grey50'

abrs_zc25 <- read_csv('data/AtlBRS_Zc_smoothSDE_data_25Hz.csv',
                    show_col_types = FALSE) 
abrs_zc25 <- left_join(abrs_zc25, CEE_meta, by = 'whaleID') |>
  mutate(cee_status = case_when( time_hr < cee_start_hr ~ 'pre',
                                 time_hr >= cee_start_hr & time_hr <= cee_end_hr ~ 'during',
                                 time_hr > cee_end_hr ~ 'post'),
         cee_status = fct_relevel(cee_status, 'pre', 'during', 'post'),
         msa1 = msa1 * 9.81) |>
  # 4 hour adj is for local time from UTC
  mutate(datetime = lubridate::seconds(time_hr * 3600) + tagon_time - lubridate::hours(4)) 

# put a binary variable clicking/not into the time series data for plotting
abrs_zc25 <- mutate(abrs_zc25, clicking = 0, foraging_period = 0)
for (d in c(1:nrow(echolocation))){
  abrs_zc25 <- abrs_zc25 |>
    mutate(clicking = if_else(whaleID == pull(echolocation, whaleID)[d] & 
                                  time_hr >= pull(echolocation, soc_hr)[d] & 
                                  time_hr <= pull(echolocation, eoc_hr)[d],
                              1,
                              clicking),
           foraging_period = if_else(whaleID == pull(echolocation, whaleID)[d] & 
                                  time_hr >= pull(echolocation, soc_hr)[d] & 
                                  time_hr <= pull(echolocation, eoc_hr)[d],
                              d,
                              foraging_period)) 
}

abrs_zc25 <- mutate(abrs_zc25,
                  soc = if_else(clicking - lag(clicking) == 1, 1, 0),
                  eoc = if_else(clicking - lag(clicking) == -1, 1, 0))


if (!file.exists('data/abrs_zc_dtag_25Hz.csv')){
write_csv(abrs_zc25, file = 'data/abrs_zc_dtag_25Hz.csv')
}

if (!file.exists('data/abrs_zc_msa_25Hz.csv')){
write_csv(abrs_zc25 |> select(whaleID, time_hr, msa1, datetime, cee_status), 
          file = 'data/abrs_zc_msa_25Hz.csv')
}


abrs_zc <- abrs_zc |>
  # 4 hour adj is for local time from UTC
  mutate(datetime = lubridate::seconds(time_hr * 3600) + tagon_time - lubridate::hours(4))
  
zc19_218a_dive_RL <- gf_line(depth ~ datetime, 
        color = ~cee_status,
        data = abrs_zc |> filter(whaleID == 'zc19_218a')) |>
  gf_point(depth ~ datetime,
        color = ~cee_status,
        data = abrs_zc |> filter(whaleID == 'zc19_218a' &
                                        (soc == 1 | eoc == 1)),
        size = 2) |>
  gf_theme(scale_color_manual('CEE', values = colrz, 
                              guide = guide_legend(nrow = 1))) |>
  gf_lims(y = c(1800, 0)) |>
  gf_labs(x = '', y = 'Depth\n(m)') |>
  gf_theme(legend.position =  'none') # c(0.75, 0.1)) 

zc19_218a_head <- gf_point(heading ~ datetime, 
                    color = ~ cee_status,
                    size = 0.5, alpha = 0.5,
                    data = abrs_zc |> filter(whaleID == 'zc19_218a')) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_labs(x = '', y = 'Heading\n(degrees)') |> 
  gf_theme(legend.position = 'none')  

this_25 <- abrs_zc25 |> filter(whaleID == 'zc19_218a' & 
                                 time_hr <= max(pull(abrs_zc |> filter(whaleID == 'zc19_218a'),
                                                     time_hr)) &
                                 time_hr >= min(pull(abrs_zc |> filter(whaleID == 'zc19_218a'),
                                                     time_hr)))

zc19_218a_head_jk <- gf_line(mag_jerk ~ datetime, 
                    color = ~ cee_status,
                    data = this_25) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_labs(x = '', y = 'Turn Angle\n(radians)') |> 
  gf_theme(legend.position = 'none')

# show the 25 Hz data and not the 1/3Hz  
zc19_218a_msa <- gf_line(msa1 ~ datetime,
                   color = ~cee_status,
                   data = this_25) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_labs(x = '', y = 'MSA\n(m/sec/sec)') |> 
  gf_theme(legend.position = 'none')
```

```{r, diffsm, fig.width = 4, fig.height = 7}

zc19_218a_depth_diffsmooth <-
  deep_ddeptht_sde$plot_par("diveprop", 
                            par_names = "sigma", 
                            n_post = 1e4,
                            covs = list(cee_dive_status = "zc19_218aCEE"), 
                            term = "cee_dive_status",
                            resp = FALSE, 
                            show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = '', 
          y = 'Depth Rate\nDifference Smooth', 
          x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)

temp <- ggplot_build(zc19_218a_depth_diffsmooth)
temp$data[[1]]$fill = 'grey70'
temp$plot$theme$text$size = 16
zc19_218a_depth_diffsmooth <- ggplot_gtable(temp)
```

```{r, zc19_218a-tagdata-diffsmooth, fig.width = 12, fig.height = 7}

tag_data <- plot_grid(zc19_218a_dive_RL, zc19_218a_head, zc19_218a_msa,
                      rel_heights = c(1.5,1,1),
                      ncol = 1,
                      align = 'hv')

# diff_smooths <- plot_grid(p4_depth_diffsmooth)

plot_grid(tag_data, zc19_218a_depth_diffsmooth,
          ncol = 2,
          rel_widths = c(2,1))
```

```{r, zc17_234a-panels, fig.width = 7.5, fig.height = 3}
w <- 'zc17_234a'
  
zc17_234a_dive_RL <- gf_line(depth ~ datetime, 
        color = ~cee_status,
        data = abrs_zc |> filter(whaleID == w)) |>
  gf_point(depth ~ datetime, 
        color = ~cee_status,
        data = abrs_zc |> filter(whaleID == w & 
                                        (soc == 1 | eoc == 1)),
        size = 2) |>
  gf_theme(scale_color_manual('CEE', values = colrz,
                              guide = guide_legend(nrow = 1))) |>
  gf_lims(y = c(1800, 0)) |>
  gf_labs(x = '', y = 'Depth (m)') |>
  gf_theme(legend.position =  'none') # c(0.75, 0.1)) 

zc17_234a_head <- gf_point(heading ~ datetime, 
                    color = ~ cee_status,
                    size = 0.5, alpha = 0.5,
                    data = abrs_zc |> filter(whaleID == w)) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_labs(x = '', y = 'Heading\n(degrees)') |> 
  gf_theme(legend.position = 'none')  

this_25 <- abrs_zc25 |> filter(whaleID == w & 
                                 time_hr <= max(pull(abrs_zc |> filter(whaleID == w),
                                                     time_hr)) &
                                 time_hr >= min(pull(abrs_zc |> filter(whaleID == w),
                                                     time_hr)))

zc17_234a_head_jk <- gf_line(mag_jerk ~ datetime, 
                    color = ~ cee_status,
                    data = this_25 |> filter(whaleID == w)) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_labs(x = '', y = 'Turn Angle\n(radians)') |> 
  gf_theme(legend.position = 'none')

# show the 25 Hz data not the 1/3Hz
zc17_234a_msa <- gf_line(msa1 ~ datetime,
                   color = ~cee_status,
                   data = this_25 |> filter(whaleID == w)) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_labs(x = '', y = 'MSA\n(m/sec/sec)') |> 
  gf_theme(legend.position = 'none')


zc17_234a_depth_diffsmooth <-
 shallow_ddepth_sde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = paste0(w, 'CEE')), 
              term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = '', 
    y = 'Depth Rate\nDifference Smooth', 
    x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)

temp <- ggplot_build(zc17_234a_depth_diffsmooth)
temp$data[[1]]$fill = 'grey70'
temp$plot$theme$text$size = 16
zc17_234a_depth_diffsmooth <- ggplot_gtable(temp)
```

```{r, zc17_234a-tagdata-diffsmooth, fig.width = 12, fig.height = 7}

tag_data <- plot_grid(zc17_234a_dive_RL, 
                      zc17_234a_head,
                      zc17_234a_msa,
                      rel_heights = c(1.5,1,1),
                      ncol = 1,
                      align = 'hv')

plot_grid(tag_data, zc17_234a_depth_diffsmooth,
          ncol = 2,
          rel_widths = c(2,1))
```

```{r, zc20_232a-panels, fig.width = 7.5, fig.height = 3}
w <- 'zc20_232a'
  
zc20_232a_dive_RL <- gf_line(depth ~ datetime, 
        color = ~cee_status,
        data = abrs_zc |> filter(whaleID == w)) |>
  gf_point(depth ~ datetime, 
        color = ~cee_status,
        data = abrs_zc |> filter(whaleID == w & 
                                        (soc == 1 | eoc == 1)),
        size = 2) |>
  gf_theme(scale_color_manual('CEE', values = colrz,
                              guide = guide_legend(nrow = 1))) |>
  gf_lims(y = c(1800, 0)) |>
  gf_labs(x = '', y = 'Depth (m)') |>
  gf_theme(legend.position =  'none') # c(0.75, 0.1)) 

zc20_232a_head <- gf_point(heading ~ datetime, 
                    color = ~ cee_status,
                    data = abrs_zc |> filter(whaleID == w),
                    size = 0.5, alpha = 0.5) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_labs(x = '', y = 'Heading\n(degrees)') |> 
  gf_theme(legend.position = 'none')  

this_25 <- abrs_zc25 |> filter(whaleID == w & 
                                 time_hr <= max(pull(abrs_zc |> filter(whaleID == w),
                                                     time_hr)) &
                                 time_hr >= min(pull(abrs_zc |> filter(whaleID == w),
                                                     time_hr)))

zc20_232a_head_jk <- gf_line(mag_jerk ~ datetime, 
                    color = ~ cee_status,
                    data = this_25 |> filter(whaleID == w)) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_labs(x = '', y = 'Turn Angle\n(radians)') |> 
  gf_theme(legend.position = 'none')

# show the 25 Hz data not the 1/3Hz
zc20_232a_msa <- gf_line(msa1 ~ datetime,
                   color = ~cee_status,
                   data = this_25 |> filter(whaleID == w)) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_labs(x = '', y = 'MSA\n(m/sec/sec)') |> 
  gf_theme(legend.position = 'none')


zc20_232a_depth_diffsmooth1 <-
 shallow_ddepth_sde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = paste0(w, 'CEE')), 
              term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = '', 
    y = 'Depth Rate\nDifference Smooth', 
    x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)

temp <- ggplot_build(zc20_232a_depth_diffsmooth1)
temp$data[[1]]$fill = 'grey70'
temp$plot$theme$text$size = 16
zc20_232a_depth_diffsmooth1 <- ggplot_gtable(temp)

zc20_232a_depth_diffsmooth2 <- 
  deep_ddeptht_sde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = paste0(w, 'CEE')), term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = '', 
          y = 'Depth Rate\nDifference Smooth', 
          x = 'Proportion of Dive') |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)

temp <- ggplot_build(zc20_232a_depth_diffsmooth2)
temp$data[[1]]$fill = 'grey70'
temp$plot$theme$text$size = 16
zc20_232a_depth_diffsmooth2 <- ggplot_gtable(temp)
```

```{r, zc20_232a-tagdata-diffsmooth, fig.width = 12, fig.height = 7}

tag_data <- plot_grid(zc20_232a_dive_RL, 
                      zc20_232a_head,
                      zc20_232a_msa,
                      rel_heights = c(1.5,1,1),
                      ncol = 1,
                      align = 'hv')

diff_smooths <- plot_grid(zc20_232a_depth_diffsmooth1,
                          zc20_232a_depth_diffsmooth2,
                          rel_heights = c(1,1),
                          ncol = 1,
                          align = 'hv')

plot_grid(tag_data, diff_smooths,
          ncol = 2,
          rel_widths = c(2,1))
```

```{r, dive-profiles, fig.width=6, fig.height = 2.5}
zc20_232a_dive_RL |>
  gf_labs(x = 'Local Time')

zc17_234a_dive_RL |>
  gf_labs(x = 'Local Time')

zc19_218a_dive_RL |>
  gf_labs(x = 'Local Time')
```

```{r, heading, fig.width=6, fig.height=1.75}
zc20_232a_head |>
  gf_labs(x = 'Local Time')

zc17_234a_head |>
  gf_labs(x = 'Local Time')

zc19_218a_head |>
  gf_labs(x = 'Local Time')
```

```{r, msa, fig.width=6, fig.height=1.75}
zc20_232a_msa |>
  gf_labs(x = 'Local Time')

zc17_234a_msa |>
  gf_labs(x = 'Local Time')

zc19_218a_msa |>
  gf_labs(x = 'Local Time')
```

Version with overlaid dive panel

```{r, overlaid-deep}
this_overlaid_deep <- abrs_zc_deep |>
          filter(whaleID == 'zc20_232a' | cee_dive_status == 'pre or post') |> 
          mutate(cee_dive_status = factor(cee_dive_status))

zc20_232a_overlaid_deep <- gf_line(depth ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = this_overlaid_deep) |>
  gf_point(depth ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = this_overlaid_deep |> filter(soc == 1 | eoc == 1),
        size = 2) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_lims(y = c(1800, 0)) |>
  gf_labs(x = '', # # x = 'Proportion total dive time',
    y = 'Depth (m)',
          title = 'Deep Dives') |>
  gf_theme(legend.position = 'none')

this_overlaid_deep <- abrs_zc_deep |>
          filter(whaleID == 'zc19_218a' | cee_dive_status == 'pre or post') |> 
          mutate(cee_dive_status = factor(cee_dive_status))

zc19_218a_overlaid_deep <- gf_line(depth ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = this_overlaid_deep) |>
  gf_point(depth ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = this_overlaid_deep |> filter(soc == 1 | eoc == 1),
        size = 2) |> 
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_lims(y = c(1800, 0)) |>
  gf_labs(x = '', # # x = 'Proportion total dive time', 
    y = 'Depth (m)',
          title = 'Deep Dives') |>
  gf_theme(legend.position = 'none')
```


```{r, overlaid-shallow, fig.height=7.5, fig.width = 6.5, warning = FALSE}
zc17_234a_overlaid_shallow <- gf_line(depth ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = abrs_zc_shallow |> filter(cee_dive_status %in% c('zc17_234aCEE', 'pre or post'))) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_lims(y = c(450, 0)) |>
  gf_labs(x = '', # # x = 'Proportion total dive time', 
    y = 'Depth (m)',
          title = 'Shallow Dives')  |>
  gf_theme(legend.position = 'none')

zc20_232a_overlaid_shallow <- gf_line(depth ~ diveprop, 
        color = ~cee_dive_status,
        group = ~ID,
        data = abrs_zc_shallow |> filter(cee_dive_status %in% c('zc20_232aCEE', 'pre or post'))) |>
  gf_theme(scale_color_manual('CEE', values = colrz)) |>
  gf_lims(y = c(450, 0)) |>
  gf_labs(x = '', # 'Proportion total dive time', 
          y = 'Depth (m)',
          title = 'Shallow Dives') |>
  gf_theme(legend.position = 'none')
```



```{r, zc20_232a-tagdata-diffsmooth2, fig.width = 12, fig.height = 7}

tag_data <- plot_grid(zc20_232a_dive_RL, 
                      zc20_232a_head,
                      zc20_232a_msa,
                      rel_heights = c(1.5,1,1),
                      ncol = 1,
                      align = 'hv')

diff_smooths <- plot_grid(zc20_232a_overlaid_deep,
                          zc20_232a_overlaid_shallow,
                          zc20_232a_depth_diffsmooth2,
                          zc20_232a_depth_diffsmooth1,
                          rel_heights = c(1,1),
                          rel_widths = c(1,1),
                          ncol = 2, nrow = 2,
                          align = 'hv')

plot_grid(tag_data, diff_smooths,
          ncol = 2,
          rel_widths = c(5,4))
```



```{r, zc19_218a-tagdata-diffsmooth2, fig.width = 12, fig.height = 7}

tag_data <- plot_grid(zc19_218a_dive_RL, 
                      zc19_218a_head,
                      zc19_218a_msa,
                      rel_heights = c(1.5,1,1),
                      ncol = 1,
                      align = 'hv')

diff_smooths <- plot_grid(zc19_218a_overlaid_deep,
                          zc19_218a_depth_diffsmooth,
                          rel_heights = c(1,1),
                          ncol = 1, nrow = 2,
                          align = 'hv')

plot_grid(tag_data, diff_smooths,
          ncol = 2,
          rel_widths = c(2,1))
```

```{r, zc17_234a-tagdata-diffsmooth2, fig.width = 12, fig.height = 7}

tag_data <- plot_grid(zc17_234a_dive_RL, 
                      zc17_234a_head,
                      zc17_234a_msa,
                      rel_heights = c(1.5,1,1),
                      ncol = 1,
                      align = 'hv')

diff_smooths <- plot_grid(zc17_234a_overlaid_shallow,
                          zc17_234a_depth_diffsmooth,
                          rel_heights = c(1,1),
                          ncol = 1, nrow = 2,
                          align = 'hv')

plot_grid(tag_data, diff_smooths,
          ncol = 2,
          rel_widths = c(2,1))
```

# Graphs Version 2

- Convert time from UTC to local (-4 hours)
- Add actual RLs from file
- all whale data together: one col per whale, several rows each

```{r, rl-data-in}

rlfiles <- tibble(whaleID = c('zc17_234a',
                              'zc19_218a',
                              'zc20_232a'),
                  rlfile = c('data/Zc17_234a RLs.csv',
                             'data/Zc19_218a RLs.csv',
                             'data/Zc20_232a_RLs.csv'))
wids <- unique(rlfiles$whaleID)
audit_data <- list()
rl_data <- list()
for (w in c(1:length(wids))){
  this_afile <- paste0('data/audit/', wids[w], 'aud.txt')
  this_rlfile <- rlfiles |> filter(whaleID == wids[w]) |> pull('rlfile')
  these_rls <- read_csv(this_rlfile, show_col_types = FALSE, na = c('', ' ', "NA", "NaN")) 
  if (file.exists(this_afile)){
  audit_data[[w]] <- read_delim(this_afile,
                           delim = '\t',
                           col_names = c('time', 'dur', 'comment')) |>
    mutate(time = parse_number(time)) |>
    filter(tolower(comment) == 'cee')
  }else{
    audit_data[[w]] <- data.frame(time = NA, dur = NA, comment = NA)
  }
  if (nrow(audit_data[[w]]) == nrow(these_rls)){
    rl_data[[w]] <- bind_cols(these_rls, audit_data[[w]])
  }else{
    if(nrow(these_rls) < nrow(audit_data[[w]])){
      n_na_pad <- nrow(audit_data[[w]]) - nrow(these_rls)
      these_rls <- bind_rows(NA * these_rls[1:n_na_pad,],
                             these_rls)
                       
      rl_data[[w]] <- bind_cols(these_rls, audit_data[[w]])
    }else{
      rl_data[[w]] <- these_rls
    }
  }
  rl_data[[w]] <- mutate(rl_data[[w]], whaleID = wids[w])
}

all_RLs <- bind_rows(rl_data) |>
  select("SPL_rms",
         "SPL_Peak",
         "SNR",
         "whaleID",
         "time")

all_RLs <- left_join(all_RLs, CEE_meta, by = 'whaleID') |>
  mutate(cee_end_local = cee_end_time - lubridate::hours(4)) 

all_RLs <- all_RLs |>
  group_by(whaleID) |>
  mutate(ping_num = c(n() : 1)) |>
  ungroup() 


 all_RLs <- all_RLs |>
   mutate(ping_time_local = if_else(is.na(time),
                             # if no ping time from audit, guess using 25s ping interval and CEE nominal end time
                             cee_end_local - lubridate::seconds(25 * (ping_num - 1)),
                             # if ping times from audit are there, use them
                             tagon_time + lubridate::seconds(time) - lubridate::hours(4)),
          zRL = (160 - SPL_rms) / 60 * 1000 + 100) |>
   filter(SPL_rms > 50)
```

second axis plots 

```{r, add-rl-dots}
zc19_218a_dive_RL <- zc19_218a_dive_RL +
  lims(y = c(1800, 0)) +
  geom_point(data = all_RLs |> filter(whaleID == 'zc19_218a'), 
             mapping = aes(x = ping_time_local, y = zRL),
             inherit.aes = FALSE,
             color = colrz[2]) + 
  scale_y_reverse(
    "Depth (m)", 
    # limits = c(1500,0), # can't set manual limits on double axis (?)
    sec.axis = sec_axis(~ -1 * ((. - 100) * 60 / 1000 - 160), 
                        name = "RL (dB SPL)"),
  ) 

zc17_234a_dive_RL <- zc17_234a_dive_RL +
  lims(y = c(1800, 0)) +
  geom_point(data = all_RLs |> filter(whaleID == 'zc17_234a'), 
             mapping = aes(x = ping_time_local, y = zRL),
             inherit.aes = FALSE,
             color = colrz[2]) + 
  scale_y_reverse(
    "Depth (m)", 
    # limits = c(1500,0), # can't set manual limits on double axis (?)
    sec.axis = sec_axis(~ -1 * ((. - 100) * 60 / 1000 - 160), 
                        name = "RL (dB SPL)"),
  ) 

zc20_232a_dive_RL <- zc20_232a_dive_RL +
  lims(y = c(1800, 0)) +
  geom_point(data = all_RLs |> filter(whaleID == 'zc20_232a'), 
             mapping = aes(x = ping_time_local, y = zRL),
             inherit.aes = FALSE,
             color = colrz[2]) + 
  scale_y_reverse(
    "Depth (m)", 
    # limits = c(1500,0), # can't set manual limits on double axis (?)
    sec.axis = sec_axis(~ -1 * ((. - 100) * 60 / 1000 - 160), 
                        name = "RL (dB SPL)"),
  ) 
```

## All 3 whales

```{r, prep-timeseries-three-whales, fig.width = 20, fig.height=7}
tag_data17 <- plot_grid(zc17_234a_dive_RL, 
                      zc17_234a_head,
                      zc17_234a_msa + labs(x = 'Local Time'),
                      rel_heights = c(1.5,1,1),
                      ncol = 1,
                      align = 'hv')

tag_data19 <- plot_grid(zc19_218a_dive_RL, 
                      zc19_218a_head,
                      zc19_218a_msa+ labs(x = 'Local Time'),
                      rel_heights = c(1.5,1,1),
                      ncol = 1,
                      align = 'hv')

tag_data20 <- plot_grid(zc20_232a_dive_RL, 
                      zc20_232a_head,
                      zc20_232a_msa + labs(x = 'Local Time'),
                      rel_heights = c(1.5,1,1),
                      ncol = 1,
                      align = 'hv')
```

```{r, timeseries-three-whales, fig.width = 20, fig.height=7}
plot_grid(tag_data17, tag_data19, tag_data20,
          labels = c('A', 'B', 'C'),
          label_size = 32, vjust = 1, hjust = -1.5,
          ncol = 3,
          rel_widths = c(1,1),
          align = 'hv')
```

## All smooths
- all overlaid dive traces + smooths together, dives on left, smooths on right


```{r, prep-all-diff-smooths, fig.height=7.5, fig.width = 6.5}
theme_set(theme_bw(base_size = 11))

zc17_234a_depth_diffsmoothb <-
 shallow_ddepth_sde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = 'zc17_234aCEE'), 
              term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = '', 
    y = 'Depth Rate\nDifference Smooth',
    x = ''
    # x = 'Proportion of Dive'
    ) |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)

temp <- ggplot_build(zc17_234a_depth_diffsmoothb)
temp$data[[1]]$fill = 'grey70'
temp$plot$theme$text$size = 11
zc17_234a_depth_diffsmoothb <- ggplot_gtable(temp)

zc19_218a_depth_diffsmoothb <-
  deep_ddeptht_sde$plot_par("diveprop", 
                            par_names = "sigma", 
                            n_post = 1e4,
                            covs = list(cee_dive_status = "zc19_218aCEE"), 
                            term = "cee_dive_status",
                            resp = FALSE, 
                            show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = '', 
          y = 'Depth Rate\nDifference Smooth', 
          x = ''
          # x = 'Proportion of Dive'
          ) |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)

temp <- ggplot_build(zc19_218a_depth_diffsmoothb)
temp$data[[1]]$fill = 'grey70'
temp$plot$theme$text$size = 11
zc19_218a_depth_diffsmoothb <- ggplot_gtable(temp)

zc20_232a_depth_diffsmooth1b <-
 shallow_ddepth_sde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = 'zc20_232aCEE'), 
              term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = '', 
    y = 'Depth Rate\nDifference Smooth', 
    x = ''
    # x = 'Proportion of Dive'
    ) |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)

temp <- ggplot_build(zc20_232a_depth_diffsmooth1b)
temp$data[[1]]$fill = 'grey70'
temp$plot$theme$text$size = 11
zc20_232a_depth_diffsmooth1b <- ggplot_gtable(temp)

zc20_232a_depth_diffsmooth2b <- 
  deep_ddeptht_sde$plot_par("diveprop", par_names = "sigma", n_post = 1e4,
              covs = list(cee_dive_status = 'zc20_232aCEE'), term = "cee_dive_status",
              resp = FALSE, show_CI = "simultaneous") |>
  gf_theme(strip.text.y = element_blank()) |>
  gf_labs(title = '', 
          y = 'Depth Rate\nDifference Smooth', 
          x = 'Proportion of Dive'
          ) |>
  gf_hline(yintercept = 0, color = 'black', size = 1.1)

temp <- ggplot_build(zc20_232a_depth_diffsmooth2b)
temp$data[[1]]$fill = 'grey70'
temp$plot$theme$text$size = 11
zc20_232a_depth_diffsmooth2b <- ggplot_gtable(temp)

```

```{r, all-diff-smooths, fig.height = 10, fig.width=8}
overlaid_dives <- plot_grid(zc17_234a_overlaid_shallow + 
                              labs(title = 'zc17_234, Shallow Dives'),
                           zc19_218a_overlaid_deep + 
                             labs(title = 'zc19_218, Deep Dives'),
                           zc20_232a_overlaid_shallow + 
                             labs(title = 'zc20_232, Shallow Dives'),
                           zc20_232a_overlaid_deep + 
                             labs(title = 'zc20_232, Deep Dives',
                                   x = 'Proportion of Dive'),
                           rel_heights = c(1,1,1,1),
                           labels = c('A', 'B', 'C', 'D'),
                           label_size = 22,
                           nrow = 4,
                           align = 'hv')

diff_smooths <- plot_grid(zc17_234a_depth_diffsmoothb,
                          zc19_218a_depth_diffsmoothb,
                          zc20_232a_depth_diffsmooth1b,
                          zc20_232a_depth_diffsmooth2b,
                          rel_heights = c(1,1,1,1),
                          ncol = 1,
                          align = 'hv')

plot_grid(overlaid_dives,
          diff_smooths,
          rel_widths = c(1,1),
          ncol = 2,
          align = 'hv',
          greedy = FALSE)
```